\section{Methodology}
\label{sec:methodology}

This section describes the implementation of both data structures, the benchmark framework, and the experimental setup.

\subsection{Implementation Details}
\label{subsec:implementation}

Both data structures are implemented from scratch in Java (no external libraries), adhering to the curriculum requirements of CSD201.

\subsubsection{Hash Table with Separate Chaining (\texttt{HashTableSC})}

Our Hash Table uses an array of singly-linked lists for collision resolution. Key design decisions include:
\begin{itemize}[nosep]
    \item \textbf{Prime-sized table}: The table capacity is always set to the next prime $\geq n$, improving hash distribution under modular compression~\citep{cormen2022}.
    \item \textbf{Load factor threshold}: Dynamic resizing triggers when $\alpha > 0.75$, doubling the capacity to the next prime.
    \item \textbf{Hash function}: Java's \texttt{String.hashCode()} with bitmask (\texttt{\& 0x7FFFFFFF}) to ensure non-negative indices.
    \item \textbf{Head insertion}: New nodes are prepended to the chain in $O(1)$ time.
\end{itemize}

\subsubsection{Bloom Filter (\texttt{BloomFilter})}

The Bloom Filter uses a \texttt{boolean[]} array of $m$ bits with $k = 7$ hash functions generated via double hashing~(\Cref{eq:doublehash}):
\begin{itemize}[nosep]
    \item \textbf{Primary hash} $h_1$: Java's \texttt{String.hashCode()} with bitmask.
    \item \textbf{Secondary hash} $h_2$: FNV-1a algorithm~\citep{carter1979} with forced odd output to ensure full-period coverage.
    \item \textbf{Parameters}: Computed dynamically using \Cref{eq:optimal_m,eq:optimal_k} with target $p = 0.01$.
\end{itemize}

\Cref{alg:bf_lookup} shows the lookup procedure.

\begin{algorithm}[t]
\caption{Bloom Filter Membership Query}
\label{alg:bf_lookup}
\begin{algorithmic}[1]
\Require Key $x$, bit array $B[0..m{-}1]$, hash functions $h_1, h_2$, count $k$
\Ensure \texttt{true} (probably in set) or \texttt{false} (definitely not)
\For{$i \gets 0$ \textbf{to} $k - 1$}
    \State $\mathit{pos} \gets (h_1(x) + i \cdot h_2(x)) \bmod m$
    \If{$B[\mathit{pos}] = 0$}
        \State \Return \texttt{false} \Comment{Definitely not in set}
    \EndIf
\EndFor
\State \Return \texttt{true} \Comment{Probably in set}
\end{algorithmic}
\end{algorithm}

\subsection{Benchmark Framework}
\label{subsec:benchmark}

The benchmark framework (\texttt{BenchmarkRunner}) orchestrates the following pipeline for each combination of dataset size $N$ and memory budget $M$:

\begin{enumerate}[nosep]
    \item \textbf{Load}: Parse the JSON dataset containing $N$ unique username strings.
    \item \textbf{Insert}: Measure wall-clock time to insert all $N$ elements.
    \item \textbf{Warm-up}: Perform 2 warm-up iterations of the lookup workload to stabilize JIT compilation.
    \item \textbf{Lookup}: Time $N$ mixed queries (50\% positive, 50\% negative) and compute throughput.
    \item \textbf{FPR Measurement} (Bloom Filter only): Execute 5{,}000 guaranteed-negative queries and count false positives.
    \item \textbf{Memory}: Record estimated memory consumption via structural analysis.
\end{enumerate}

Garbage collection is forced (\texttt{System.gc()}) between runs to minimize cross-contamination.

\subsection{Experimental Configuration}
\label{subsec:config}

\Cref{tab:config} summarizes the experimental parameters. The JVM flag \texttt{-Xmx} enforces a hard heap ceiling, causing an \texttt{OutOfMemoryError} when the budget is exceeded.

\begin{table}[t]
\centering
\caption{Experimental Configuration}
\label{tab:config}
\begin{tabular}{@{}ll@{}}
\toprule
\textbf{Parameter} & \textbf{Value} \\
\midrule
Language \& Runtime      & Java (OpenJDK), single-threaded \\
Memory Budgets $M$       & 16, 32, 64, 128, 256~MB (\texttt{-Xmx}) \\
Dataset Sizes $N$        & 1K, 5K, 10K, 25K, 50K, 100K, 250K, 500K, 1M, 2M \\
Data Format              & JSON arrays of unique username strings \\
Hash Table Load Factor   & $\alpha_{\max} = 0.75$ (auto-resize) \\
Hash Table Capacity      & Next prime $\geq N$ \\
Bloom Filter Target FPR  & $p = 0.01$ (1\%) \\
Bloom Filter Hash Count  & $k = 7$ (computed from~\Cref{eq:optimal_k}) \\
Lookup Workload          & $N$ mixed queries (50\% positive / 50\% negative) \\
FPR Test Queries         & 5{,}000 guaranteed-negative strings \\
Warm-up Iterations       & 2 (for JIT stabilization) \\
\bottomrule
\end{tabular}
\end{table}

% Figure 1: System Architecture Diagram
\begin{figure}[t]
\centering
\begin{tikzpicture}[
    node distance=1.2cm and 1.5cm,
    box/.style={draw, rounded corners=3pt, minimum height=0.9cm, minimum width=2.2cm, font=\small, align=center, thick},
    arrow/.style={-{Stealth[length=3mm]}, thick}
]

% Dataset
\node[box, fill=lightgray] (data) {JSON Dataset\\$N$ usernames};

% Insert phase
\node[box, fill=htblue!20, right=1.8cm of data] (insert) {Insert Phase\\Timer};

% Fork
\node[box, fill=htblue!30, above right=0.8cm and 1.8cm of insert] (ht) {Hash Table\\(Separate Chain)};
\node[box, fill=bfgreen!30, below right=0.8cm and 1.8cm of insert] (bf) {Bloom Filter\\($k{=}7$, $m$ bits)};

% Lookup
\node[box, fill=htblue!20, right=1.8cm of ht] (htlook) {Lookup\\$N$ queries};
\node[box, fill=bfgreen!20, right=1.8cm of bf] (bflook) {Lookup\\$N$ queries};

% Metrics
\node[box, fill=warnorg!20, right=4.2cm of insert] (metrics) {Metrics\\Collector};

% Results
\node[box, fill=oomred!15, right=1.5cm of metrics] (csv) {CSV\\Output};

% Arrows
\draw[arrow] (data) -- (insert);
\draw[arrow] (insert) |- (ht);
\draw[arrow] (insert) |- (bf);
\draw[arrow] (ht) -- (htlook);
\draw[arrow] (bf) -- (bflook);
\draw[arrow] (htlook) -| (metrics);
\draw[arrow] (bflook) -| (metrics);
\draw[arrow] (metrics) -- (csv);

% Labels
\node[font=\scriptsize, above=0.05cm of ht, text=htblue] {Exact};
\node[font=\scriptsize, below=0.05cm of bf, text=bfgreen] {Probabilistic};

% Memory constraint box
\draw[dashed, thick, oomred] ([xshift=-0.3cm, yshift=0.8cm]ht.north west) rectangle ([xshift=0.3cm, yshift=-0.8cm]bf.south east);
\node[font=\scriptsize\bfseries, text=oomred, anchor=south] at ([yshift=0.85cm]$(ht.north west)!0.5!(bf.north east)$) {JVM -Xmx $M$~MB};

\end{tikzpicture}
\caption{Benchmark system architecture. Each dataset--budget combination runs the insert--lookup pipeline for both data structures under a hard JVM memory ceiling. Metrics (throughput, memory, FPR, collisions) are collected and exported to CSV.}
\label{fig:architecture}
\end{figure}
